On Optimality of Finite Element Integration

### 2016-02-26 TOMS_Paper/assembly.tex
Improves explonation of sharing elimination algo

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont 
~
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions. For all multilinear operands in $P \in \mathbb{P}$ such that $|P| < |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of Proposition~\ref{prop:multi-struct}.}

\item Build the \textit{sharing graph} $G= (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced at step (1). An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.

\item Let $T \subset S$ be the subset of vertices with at most one incident edges, or ``terminals''. Partition $T$ into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that the terminals in $T^i$ are either adjacent to the same vertex or disconnected. Apply Strategy~\ref{strategy:ii} factorizing the terminals in each $T^i$, provided that this leads to a reduction in operation count.\\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation; the cases in which the expansion cost exceeds the code motion gain are discarded.}

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ will not cause an increase in operation count.\\
\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping $[s^1, ..., s^m]$, which impact the expansion cost.}

\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}
\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (3) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (2)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$. Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.

>>>>

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont 
~
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions. For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: as a consequence of Proposition~\ref{prop:multi-struct}, $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied}

\item For each sub-expression $expr$ depending on exactly one linear loop, collect the multilinear symbols and the temporaries produced at step (1). Partition them into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that $T^i$ includes all instances of a given symbol in $expr$. Apply Strategy~\ref{strategy:ii} factorizing the symbols in each $T^i$, provided that this leads to a reduction in operation count; otherwise, apply Strategy~\ref{strategy:i} \\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation. In the cases in which expansion outweighs code motion, Strategy~\ref{strategy:i} is preferred.}\\
\textit{Note: the expansion cost is a function of the products wrapping a symbol (how many of them and their arity), so it can be determined in a tree visit.}

\item Build the \textit{sharing graph} $G = (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced by the previous steps. An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.\\
\textit{Note: the following steps will only impact bilinear forms, since otherwise $E = \emptyset$.}

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.

\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (2) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$. Since there are no more factorization opportunities, the ILP formulation becomes irrelevant.
